Altering Backward Pass Gradients improves Convergence

### 2022-09-18 paper.tex
Mod6

\documentclass[times,sort&compress]{elsarticle}
\usepackage{ycviu}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{graphicx}

@@

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}

@@

\textbf{Type 0:} No modification: In this method, we calculate the gradients using the
standard calculus rules and use the chain rule to calculate the gradients of the rest of
the network's parameters, also known as backpropagation (Fig.
\ref{fig:gradient_altering}a). The network is then updated with the gradient descent
equation:

@@

\textbf{Type I:} Gradient modification at multiple points independently: In this method,
the gradients are first computed using standard procedure and then individually altered
(Fig. \ref{fig:gradient_altering}b). Gradient clipping \cite{pascanu2013difficulty} and
Adaptive gradient clipping \cite{brock2021high} are examples of such modifications. In
both of these methods, the gradients are first computed using standard rules and then
they are modified using some function. It can be described as:

@@

\textbf{Type II:} Gradient modification at a point very early in the backward graph: In
this type of modification, the gradient is altered at a very early stage in the backward
computation graph and then all subsequent gradients are generated using the values
obtained with the modified gradients (Fig. \ref{fig:gradient_altering}c). Because of the
chain rule, network parameters whose gradients are connected to the point of alteration
in the computation graph also gets subsequently altered. It can be described as:

@@

\subsection{Experiments on Normalized Networks}
\label{sec:bn}

\begin{table}[t]
\centering
\caption{ Results and comparisons for networks trained on ImageNet-1K. All experiments
are performed with a 100 epoch budget. }
\label{tab:imagenet_table}
\vspace{5pt}
\begin{tabular}{cccccc}
\textbf{Model} & \textbf{Scheduler} & \textbf{Method} & \textbf{Train Acc} &
\textbf{Test Acc} & \textbf{Diff} \\
\midrule
ResNet-18 & Step & - & 69.95 & 69.704 & - \\
& Step & PGT & 70.3 & \textbf{69.844} & 0.14 \\
\midrule
ResNet-18 & Cosine & - & 70.38 & 70.208 & - \\
& Cosine & PGT & 70.53 & \textbf{70.298} & 0.09 \\
\midrule
ResNet-50 & Step & - & 78.99 & 75.97 & - \\
& Step & PGT & 79.56 & \textbf{76.494} & 0.524 \\
\midrule
ResNet-50 & Cosine & - & 79.18 & 76.56 & - \\
& Cosine & PGT & 79.68 & \textbf{77.216} & 0.656 \\
\midrule
ResNet-101 & Cosine & - & 82.29 & 77.896 & - \\
& Cosine & PGT & 83.1 & \textbf{78.258} & 0.362 \\
\midrule
SEResNet-18 & Cosine & - & 71.42 & 71.09 & - \\
& Cosine & PGT & 71.6 & \textbf{71.436} & 0.346 \\
\midrule
SEResNet-50 & Cosine & - & 81.5 & 77.218 & - \\
& Cosine & PGT & 82.47 & \textbf{77.952} & 0.734 \\
\end{tabular}
\end{table}

We perform experiments on various BN variants ResNets using the ImageNet-1K dataset
\cite{deng2009imagenet}. We observe a constant $0.09\%$ to $0.14\%$ rise in ResNet-18's
test accuracy. What's more interesting is that we are seeing comparable improvements in
training accuracy. PGT increases the generalization gap, as evident from this
observation. However, it allows the network to better fit the dataset. In section
\ref{sec:Abla}, we show how combining PowerGrad Transform with regularization techniques
such as label smoothing may improve network performance and simultaneously reduce the
generalization gap.

We find significantly larger improvements in experiments using ResNet-50, a $0.52\%$
increase over the step scheduling baseline and a $0.65\%$ increase over the cosine
scheduling baseline. In studies with Squeeze-and-Excitation networks, we use the cosine
scheduler. In these experiments, we find consistent improvements across the board, with
SEResNet-18 \cite{hu2018squeeze} gaining $0.34\%$ and SEResNet-50 \cite{hu2018squeeze}
gaining $0.73\%$. As we saw with ResNet-18, we find continuous improvements in training
performance because the networks are just training better and arriving to better optimas
at convergence. ResNet-101 training provides a modest advantage of $0.36\%$.

All models are trained on four V100 GPUs with a batch size of $1024$. We utilize the
same set of hyperparameters in each experiment, which are as follows: 100 epoch budget,
5 epochs linear warmup phase beginning with a learning rate of $4\times 10^{-4}$ and
ending with a peak learning rate of $0.4$. In our studies, we employ either a step
scheduler (dividing the learning rate by $10$ at the $30^{th}$, $60^{th}$, and $90^{th}$
epochs) or a cosine decay scheduler\footnote{Reproducible code, training recipes for the
experiments, pretrained checkpoints and training logs are provided at: \\
\url{https://github.com/bishshoy/power-grad-transform}. Code is adapted from
\cite{rw2019timm}.}. We observe a consistent improvement in test metrics when utilizing
cosine scheduling across all of our experiments. Finally, towards the conclusion of
training, we impose a 10 epoch cooling phase in which the network is trained with a
constant learning rate schedule set at $4\times 10^{-4}$. Other hyperparameters include
a weight decay value of $5\times 10^{-4}$ and a momentum of $0.9$ for the SGD Nesterov
optimizer. All of our models are trained using mixed floating point precision. The
findings are shown in Table \ref{tab:imagenet_table}. With these findings, we observe
that our proposed altering gradient mechanism PGT improves the model performance further
even for batch normalized networks.

\subsection{Empirical Studies on Residual Networks without Batch Normalization}
\label{sec:Empi}

This section empirically examines some of the issues that occur when networks are
trained without normalization layers. ResNet-18 \cite{he2016deep} is used as the
foundation model for all empirical trials owing to its popularity among deep learning
practitioners and the relative simplicity with which it can be trained on a big dataset
such as ImageNet \cite{russakovsky2015imagenet}. ResNet-18 is composed of $20$
convolutional layers and one fully connected (FC) layer. At the input, it has a single
convolutional layer with filters of size $7\times 7$. Three convolutional layers are
placed in the downsampling skip connections. The remaining convolutional layers have
several $3\times 3$ filters that are grouped together to create a block. Each basic
block is connected by a skip connection. At the end of all convolutional layers, a
global average pooling layer downsamples the features to a fixed-size feature vector and
transfers it to the FC layer, which fits the features using regression and outputs
logits. The logits $z_i$ produced by the FC layer are fed into a softmax layer, which
transforms them to predicted probabilities $p_i$ using the equation $p_i=\frac
{e^{z_i}}{\sum _j e^{z_j}}$.

We concentrate on the following metric: the per-filter L2-norm of each layer. Throughout
the training process, we monitor variations in the per-filter weight norm. Because a
full plot of each layer's data would use an inordinate amount of space, we focus on a
few critical layers' plots. The supplementary section contains plots for each layer.
ResNet-18 without any normalization layers is used for all experiments in this section.
The norm of the weights of the $11^\textit{th}$ convolutional layer is shown in Fig.
\ref{fig:norm_plots}a. Layer $11$ includes $256$ filters. Each colour on the figure
represents the evolution of a particular filter throughout the course of training
iterations. The critical point to note is that some filters achieve a norm of zero
during training. This event is referred to as `Zeroing Out', and occurs when a channel
of a weight tensor gets fully filled with zeros. If all the coefficients of the filter
is 0 then only L2 norm of a filter can become zero. If we observe zero filter norm for a
filter, then that filter does not contribute at all to determine the input-output
relationship of a dataset. According to our findings, when a filter tensor becomes
zeroed out, it does not recover with further training.

@@

Similar to the plot \ref{fig:norm_plots}a, we see the plot of the final convolutional
layer in Fig. \ref{fig:norm_plots}b, where we observe a number of filters / units
completely dropping out towards the end of training leading to underutilization of the
network's resources. We also plot the final layer's feature norm computed in a per
channel fashion. In Fig. \ref{fig:norm_plots}c, we see the norm of the feature vector at
the output side of the global average pooling layer of ResNet-18. The feature vector
here directly interfaces with the fully connected layer immediately after. Any zeroed
out regions in this tensor directly leads to permanent information loss, as it does not
contribute to the learning of decision boundaries in the upcoming fully connected layer.

We observe both effects \textbf{1)} zeroed-out filter \textbf{2)} zeroed-out feature.
The impact of having a zeroed-out filter or a feature might have a variety of
consequences for training and optimization. If a zeroed out filter is present in a
convolutional layer, there are two possible outcomes. For starters, if the layer is
immediately followed by a fully connected regression layer, there is an abrupt loss of
information. For the reason that a zeroed out filter always generates zeroed out
features regardless of the input image, when this tensor is fed to a fully connected
layer, it does not communicate any information that is useful for classification. In the
event that the zeroed out filter has a skip connection linked across it, this may be
circumvented as long as there are no overlapping zeroed out channels in the two feature
maps.

On the other hand, if a feature tensor has a zeroed out channel at the input of a layer,
the layer creates a zeroed out feature channel at the output, regardless of the weights.
However, if there is a zeroed out channel in the feature map just before the fully
connected layer, then this results in a permanent loss of information as it does not
contribute to the learning of decision boundaries. ResNet-18 suffers from the same kind
of information loss, as the final feature tensor (which is generated after the global
average pooling layer) has multiple channels that are zeroed out [Fig.
\ref{fig:norm_plots}c]. This information collapse phenomenon occurs when a channel that
has been zeroed out stays zeroed out for all images in the dataset.

@@

The impact of a feature tensor or weight tensor being zeroed out is seen to be
particularly common at larger batch size training procedures in networks without
normalization layers. As a result, the network's capacity is underutilized since some of
the filters are not involved in extracting any useful information that is important for
performing classification. When working with large batch sizes, it is possible that an
entire layer zeroes out [Fig. \ref{fig:high_bs}]. In a network without skip connections,
this can cause a full collapse in training. For example if a weight tensor completely
zeroes out, and it always produces a zero feature tensor, then the input feature tensor
to the next convolutional layer is a zero tensor and thus the output of the next layer
is always a zero tensor irrespective of its weights, and this continues until the end of
the neural network pipeline leading to zero logits. Also, zeroed out weights tensors
lead to zeroed out gradients hence stopping training for all subsequent iterations
leading to a collapse in training. Although residual networks bypass this difficulty via
the use of the skip connection, it does not solve the layer's non-participation in the
overall objective of classification. It is possible to solve this issue by using methods
such as gradient clipping (GC), adaptive gradient clipping (AGC), and PowerGrad
Transform (PGT), all of which regulate the amount of gradient that flows at various
junctions of the network. GC and AGC modify the gradients at each and every filter of
every layer. PGT performs this operation at only one junction, which is at the softmax
layer.

@@

In a similar fashion to the study in section \ref{sec:Empi}, we examine the norm
profiles of features and weights in the PGT and AGC. The supplementary section includes
all the detailed graphs of each layer. In Fig. \ref{fig:norm_plots}d, we show the
per-filter norm plot of the $11^\textit{th}$ convolutional layer as it progresses through
a PGT enabled training session. We observe that the number of zeroed out filters has
considerably reduced. We observe similar benefits in plot of Fig. \ref{fig:norm_plots}e,
which is the plot of the per-filter norm of the final convolutional layer. Coming to
benefits in feature norms, we observe in plot of Fig. \ref{fig:norm_plots}f, the final
feature tensor obtained with PGT enabled training do not contain any zeroed out regions,
leading us to assert that information loss is heavily mitigated as the features pass on
from the feature extracting convolutional layers to the fully connected stage for
regression.

AGC is excellent at eliminating the issue of filter and feature zeroing out. However,
AGC recovers a lower performance benefit than PGT. In the instance of PGT, we see that
it also mitigates the zeroing out phenomenon in most layers. In Fig.\ref{fig:agc_pgt}a,
we see a plot of the $15^\textit{th}$ layer of the non-BN ResNet-18 trained without PGT.
As we can see that there has been quite a few filter dropouts in the baseline. PGT
recovers most of the filters [Fig. \ref{fig:agc_pgt}b]. In some layers however, the
performance of PGT worsens. The $2\textit{nd}$ convolutional layer has more zeroed-out
filters in PGT [Fig. \ref{fig:agc_pgt}e] than in the baseline [Fig. \ref{fig:agc_pgt}d].
Despite that, PGT achieves a greater overall accuracy. This necessitates the
simultaneous use of AGC and PGT.

The advantage of combining GC with PGT is not significant. When AGC and PGT are
combined, we see a tremendous increase in accuracy of over $2.06\%$ over the baseline.
As seen in Fig. \ref{fig:agc_pgt}c and \ref{fig:agc_pgt}f, activating AGC and PGT
concurrently results in no filter dropouts in all layers while also improving the
accuracy by a significant margin. Additionally, we give comprehensive graphs of the norm
profiles of the AGC + PGT method for all layers in the supplementary section.

\subsection{Towards removing Normalization}
\label{sec:Towa}

\begin{table}[t]
\centering
\caption{ Results for non-normalized ResNets on ImageNet-1K. Each experiment is
conducted with a 100 epoch budget. }
\label{tab:wobn_table}
\vspace{5pt}
\begin{tabular}{cccccc}
\textbf{Model} & \textbf{Batch Size} & \textbf{Method} & \textbf{Train Acc} &
\textbf{Test Acc} & \textbf{Diff} \\
\midrule
ResNet-18 (non-BN) & 1024 & - & 67.19 & 64.93 & - \\
\midrule
ResNet-18 (non-BN) & 512 & - & 68.02 & 66.552 & - \\
& 512 & PGT & 69.5 & \textbf{67.236} & 0.684 \\
\midrule
ResNet-18 (non-BN) & 256 & - & 68.86 & 66.796 & - \\
& 256 & GC & 69.04 & 67.064 & 0.268 \\
& 256 & AGC & 69.06 & 67.298 & 0.502 \\
& 256 & PGT & 69.97 & \textbf{67.814} & 1.018 \\
& 256 & GC+PGT & 68.67 & 67.088 & 0.292 \\
& 256 & AGC+PGT & 70.92 & \textbf{68.856} & 2.06 \\
\end{tabular}
\end{table}

Experiments are carried out on non-Batch Normalized networks. We specifically utilize
the ResNet-18 model and train it on ImageNet-1K which has 1.28 million images divided
into 1000 classes. ImageNet-1K's test set consists of 50 images per class. Table
\ref{tab:wobn_table} shows the findings.

To begin, we use a batch size of 1024. This is only to illustrate that in unnormalized
networks, accuracy suffers significantly when batch sizes are large. Additionally, the
training process is often unstable, resulting in frequent layer dropouts (as seen in
\ref{sec:Empi}) and sometimes infinite loss values. Thus, for all subsequent tests, we
mostly use batch sizes of 512 and 256. We find $\alpha=0.25$ and $\alpha=0.05$ to be
mostly optimal for ResNet-18 and ResNet-50, though larger values such as $\alpha=0.3$
also have good performance as well. We explore different values of $\alpha$ in section
\ref{sec:Abla}. For ResNet-18, invoking PGT with $\alpha=0.25$ improves the baseline by
$0.68\%$ when the batch size is 512, and by over $1\%$ when the batch size is 256. In
comparison, the improvement obtained by GC and AGC is much less, at $0.27\%$ and
$0.5\%$, respectively.

>>>>

\documentclass[times,sort&compress]{elsarticle}
\usepackage{ycviu}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{pifont}
\usepackage{graphicx}

% \usepackage{xcolor}
% \usepackage{pagecolor}
% \pagecolor{black}
% \color{white}

@@

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}

@@

\textbf{Type 0:} No modification: In this method, we calculate the gradients using the
standard calculus rules and use the chain rule to calculate the gradients of the rest of
the network's parameters, also known as backpropagation as portrayed in Fig.
\ref{fig:gradient_altering}(a). The network is then updated with the gradient descent
equation:

@@

\textbf{Type I:} Gradient modification at multiple points independently: In this method,
the gradients are first computed using standard procedure and then individually altered
as shown in Fig. \ref{fig:gradient_altering}(b). Gradient clipping
\cite{pascanu2013difficulty} and Adaptive gradient clipping \cite{brock2021high} are
examples of such modifications. In both of these methods, the gradients are first
computed using standard rules and then they are modified using some function. It can be
described as:

@@

\textbf{Type II:} Gradient modification at a point very early in the backward graph: In
this type of modification, the gradient is altered at a very early stage in the backward
computation graph and then all subsequent gradients are generated using the values
obtained with the modified gradients. We illustrate this type of modification in Fig.
\ref{fig:gradient_altering}(c). Because of the chain rule, network parameters whose
gradients are connected to the point of alteration in the computation graph also gets
subsequently altered. It can be described as:

@@

\subsection{Experiments on convolutional neural networks (with batch-normalization)}
\label{sec:bn}

\begin{figure}[!t]
\centering
\begin{subfigure}{.5\columnwidth}
\centering
\includegraphics[width=0.9\columnwidth]{acc_vs_epoch_r18}
\caption{Plot of training and test accuracies (ResNet-18)}
\end{subfigure}%
\begin{subfigure}{.5\columnwidth}
\centering
\includegraphics[width=0.9\columnwidth]{acc_vs_epoch_r50}
\caption{Plot of training and test accuracies (ResNet-50)}
\end{subfigure}
\caption{ Log-linear plots of training and test accuracies and comparison with baseline
of batch-normalized variants: \textbf{(a)} ResNet-18 ($\alpha=0.25$), \textbf{(b)}
ResNet-50 ($\alpha=0.3$). For better visualization of accuracies obtained in the final
iterations, we plot the y-axis in log scale. PowerGrad Transform improves upon the
baseline in all cases. The corresponding accuracies are mentioned in table
\ref{tab:imagenet_table}. }
\label{fig:metrics}
\end{figure}

\begin{table}[!t]
\centering
\caption{ Results and comparison table for CNN networks trained on ImageNet-1K (sorted
in descending order w.r.t. the last column denoting test accuracy improvements over
baseline). Best training and test accuracies are highlighted in red and blue
respectively. Accuracy differences are highlighted in yellow. In all experiments, we
observe that PGT positively improves both training and test accuracies. We also observe
that deeper networks such as SE-ResNet-50 and ResNet-50 accrue higher gains over the
baselines as compared to shallow networks like ResNet-18. }
\label{tab:imagenet_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Scheduler}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PowerGrad} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{Transform ($\alpha$)} & \textbf{Accuracy (\%)} & \textbf{Diff (\%)} &
\textbf{Accuracy (\%)} & \textbf{Diff (\%)} \\
\midrule
\multirow{2}{*}{SE-ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 81.5 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.97}}} & 77.218 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.734}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{82.47}} & & \textcolor{blue}{\textbf{77.952}} & \\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 79.18 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.5}}} & 76.56 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.656}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.68}} & & \textcolor{blue}{\textbf{77.216}} & \\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Step} & Baseline & - & 78.99 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.57}}} & 75.97 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.524}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.56}} & & \textcolor{blue}{\textbf{76.494}} & \\
\midrule
\multirow{2}{*}{ResNet-101} & \multirow{2}{*}{Cosine} & Baseline & - & 82.29 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.81}}} & 77.896 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.362}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{83.1}} & & \textcolor{blue}{\textbf{78.258}} & \\
\midrule
\multirow{2}{*}{SE-ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 71.42 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.18}}} & 71.09 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.346}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{71.6}} & & \textcolor{blue}{\textbf{71.436}} & \\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Step} & Baseline & - & 69.95 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.35}}} & 69.704 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.14}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.3}} & & \textcolor{blue}{\textbf{69.844}} & \\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 70.38 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.15}}} & 70.208 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.09}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.53}} & & \textcolor{blue}{\textbf{70.298}} & \\
\end{tabular}
\end{table}

We perform experiments on different variants ResNets using the ImageNet-1K dataset
\cite{deng2009imagenet}. All models are trained on four V100 GPUs with a batch size of
$1024$. We utilize a common set of hyperparameters for all experiments, which are as
follows: 100 epoch budget, 5 epochs linear warmup phase beginning with a learning rate
of $4\times 10^{-4}$ and ending with a peak learning rate of $0.4$, a momentum of $0.9$
and weight decay of $5\times 10^{-4}$, the SGD Nesterov optimizer and mixed precision.
In our studies, we employ either a step scheduler (dividing the learning rate by $10$ at
the $30^{th}$, $60^{th}$, and $90^{th}$ epochs) or a cosine decay scheduler
\cite{loshchilov2016sgdr}. We find $\alpha=0.25$ and $\alpha=0.05$ to be good choices
for ResNet-18 and ResNet-50, though larger values such as $\alpha=0.3$ also have good
performance as well. The experimental results are shown in table
\ref{tab:imagenet_table}, and we mention the value of the PGT hyperparmeter ($\alpha$)
in each experiment. We explore different values of $\alpha$ in section \ref{sec:Abla}
and provide detailed grid plots in the supplementary section.

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. These improvements are
obtained with the number of epochs being kept the same as that of the baseline model.
Alternatively, practioners can choose to employ PGT and have the networks converge to
accuracies similar to that of the baseline with a much lower epoch budget. We explore
this in detail in section \ref{sec:speed}. Therefore, PGT also speeds up training. In
our experiments with ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance
enhancement over the baseline while using the cosine scheduler. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ and $0.524\%$
respectively for training and testing. ResNet-101 sees a higher improvement in training
fit, $0.81\%$ to be exact, while the improvement over the test set is $0.362\%$. Smaller
networks such as SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but
nevertheless positive. Therefore we conclude that with PGT, the consistent improvements
in training accuracies across all cases is because the networks train better and arrive
at better optimas during convergence. Per epoch training and test accuracy plots of
ResNet-18 and ResNet-50 (both with and without PGT) are shown in Fig.
\ref{fig:metrics}(a, b). For better visualization of the top end of the plots, we scale
the y-axis in log scale. Our experiments with batch-normalized networks indicate that
PowerGrad Transform speeds up training and allows the network to learn better
representations\footnote{Reproducible code, training recipes for the experiments,
pretrained checkpoints and training logs are provided at: \\
\url{ https://github.com/bishshoy/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

\subsection{Empirical studies on networks without Batch Normalization}
\label{sec:Empi}

This section empirically examines some of the issues that occur when networks are
trained without normalization layers and we provide insight into the effects of
PowerGrad transform on such networks. We use ResNet-18 \cite{he2016deep} as the
foundation model for all empirical trials owing to its popularity among deep learning
practitioners and the relative simplicity with which it can be trained on a big dataset
such as ImageNet-1K \cite{russakovsky2015imagenet}. We found deeper networks (such as
ResNet-34 and ResNet-50 models) are impossible to train without BatchNormalization
because of increased depth. So in order to conduct our empirical experiments, we solely
focus on the ResNet-18 architecture. For our experiments, we designate different layers
with their corresponding layer indices. We provide a detailed layer-wise index list of
ResNet-18 in the supplementary section.

We concentrate on the following metric: the per-filter L2-norm of each layer's weight
tensor. Throughout the training process, we monitor variations in the per-filter weight
norm. Because a full plot of each layer's data would use an inordinate amount of space,
we focus on a few critical layers' plots. The supplementary section contains full plots
for all layer for various settings. The norm of the weights of the $11^\textit{th}$
convolutional layer is shown in Fig. \ref{fig:norm_plots}(a). Layer $11$ includes $256$
filters. Each colour on the figure represents the iteration-wise evolution of a
particular filter throughout the course of training. The critical point to note is that
some filters achieve a norm of zero during training. We refer to this event as `Zeroing
Out', and occurs when a channel of a weight tensor gets fully filled with zeros. If all
the coefficients of the filter becomes zero, only then the L2 norm of that filter
becomes zero. If we observe zero filter norm for a filter, then that filter does not
contribute at all to determine the input-output relationship of a dataset, as the
feature tensor it produces is also an all-zero tensor. According to our findings, when a
filter tensor becomes zeroed out, it does not recover with further training. We observe
both effects \textbf{1)} zeroed out filter \textbf{2)} zeroed out feature.

@@

Similar to the plot \ref{fig:norm_plots}(a), we see the plot of the final convolutional
layer in Fig. \ref{fig:norm_plots}(b), where we observe a number of filters / units
completely dropping out towards the end of training leading to underutilization of the
network's parameters. In Fig. \ref{fig:norm_plots}(c), we see the norm of the feature
vector at the output side of the global average pooling layer of ResNet-18. The feature
vector here directly interfaces with the fully connected layer immediately after. Any
zeroed out regions in this tensor directly leads to permanent information loss, as it
does not contribute to the learning of decision boundaries in the upcoming fully
connected layer, and indeed there exists a few zeroed out features in the baseline
model, which we discuss next. The impact of having a zeroed out filter or a feature
might have a variety of consequences for training and optimization. If a zeroed out
filter is present in a convolutional layer, there are two possible outcomes. If the
layer is immediately followed by a fully connected regression layer, there is an abrupt
loss of information. For the reason that a zeroed out filter always generates zeroed out
features regardless of the input image, when this tensor is fed to a fully connected
layer, it does not communicate any information that is useful for classification. In the
event that the zeroed out filter has a skip connection linked across it, this may be
circumvented as long as there are no overlapping zeroed out channels in the two feature
maps. On the other hand, if a feature tensor has a zeroed out channel at the input of a
layer, the layer creates a zeroed out feature channel at the output, regardless of the
weights. However, if there is a zeroed out channel in the feature map just before the
fully connected layer, then this results in a permanent loss of information as it does
not contribute to the learning of decision boundaries. ResNet-18 suffers from the same
kind of information loss, as the final feature tensor (which is generated after the
global average pooling layer) has multiple channels that are zeroed out [Fig.
\ref{fig:norm_plots}(c)]. This information collapse phenomenon occurs when a channel
that has been zeroed out stays zeroed out for all images in the dataset.

@@

The impact of a feature tensor or weight tensor being zeroed out is seen to be
particularly common in training procedures involving larger batch sizes in networks
without normalization layers. As a result, the network's parameters are underutilized
since some of the filters are not involved in extracting any useful information that is
important for performing classification. When working with large batch sizes, it is
possible that an entire layer zeroes out [Fig. \ref{fig:high_bs}]. In a network without
skip connections, this can cause a full collapse in training. For example, if a weight
tensor completely zeroes out, and it always produces a zero feature tensor, then the
input feature tensor to the next convolutional layer is a zero tensor and thus the
output of the next layer is always a zero tensor irrespective of its weights, and this
continues until the end of the neural network pipeline leading to zero logits. Also,
zeroed out weights tensors lead to zeroed out gradients hence stopping training for all
subsequent iterations leading to a collapse in training. Although residual networks
bypass this difficulty via the use of the skip connection, it does not solve the layer's
non-participation in the overall objective of classification. It is possible to solve
this issue by using methods such as gradient clipping (GC), adaptive gradient clipping
(AGC), and PowerGrad Transform (PGT), all of which regulate the amount of gradient that
flows at various junctions of the network. GC and AGC modify the gradients at each and
every filter of every layer. PGT performs this operation at only one junction, which is
at the softmax layer.

@@

In a similar fashion to the study in section \ref{sec:Empi}, we examine the norm
profiles of features and weights in the PGT and AGC. The supplementary section includes
all the detailed graphs of all layer. In Fig. \ref{fig:norm_plots}(d), we show the
per-filter norm plot of the $11^\textit{th}$ convolutional layer as it progresses
through a PGT enabled training session. We observe that the number of zeroed out filters
has considerably reduced. We observe similar benefits in plot of Fig.
\ref{fig:norm_plots}(e), which is the plot of the per-filter norm of the final
convolutional layer. Coming to benefits in feature norms, we observe in plot of Fig.
\ref{fig:norm_plots}(f), the final feature tensor obtained with PGT enabled training do
not contain any zeroed out regions, leading us to assert that information loss is
heavily mitigated as the features pass on from the feature extracting convolutional
layers to the fully connected stage for regression.

AGC is excellent at eliminating the issue of filter and feature zeroing out. However,
AGC recovers a lower performance benefit than PGT. In the instance of PGT, we see that
it also mitigates the zeroing out phenomenon in most layers. In
Fig.\ref{fig:agc_pgt}(a), we see a plot of the $15^\textit{th}$ layer of the non-BN
ResNet-18 trained without PGT. As we can see that there has been quite a few filter
dropouts in the baseline. PGT recovers most of the filters [Fig. \ref{fig:agc_pgt}(b)].
In some layers however, the performance of PGT is slightly weaker than the baseline. The
$2^\textit{nd}$ convolutional layer has more zeroed out filters in PGT [Fig.
\ref{fig:agc_pgt}(e)] than in the baseline [Fig. \ref{fig:agc_pgt}(d)]. Despite that,
PGT achieves a higher overall accuracy (table \ref{tab:wobn_table}). This necessitates
the simultaneous use of AGC and PGT, for the non-BN ResNet-18.

\begin{table}[!t]
\centering
\caption{ Results for non-normalized ResNets on ImageNet-1K. Each experiment is
conducted with a 100 epoch budget using the cosine decay scheduler. Best training and
test accuracies are highlighted in red and blue respectively. Top differences in
training and test accuracies are marked in yellow. The baselines in each segment is
taken as the reference point for all `Diff' accuracies in that segment. The first three
segments highlight comparison for experiments where the methods are applied
individually, while the last segment denotes performance metrics for combinations of
different methods (GC + PGT, AGC + PGT). We see that at any batch size, with PGT, the
performance of non-normalized version of ResNet-18 improves significantly over the
baseline owing to the improvements in training as observed in Fig. \ref{fig:agc_pgt}. }
\label{tab:wobn_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Batch Size}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PowerGrad} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{Transform ($\alpha$)} & \textbf{Accuracy (\%)} & \textbf{Diff (\%)} &
\textbf{Accuracy (\%)} & \textbf{Diff (\%)} \\
\midrule
ResNet-18 (non-BN) & 1024 & Baseline & - & 67.19 & - & 64.93 & - \\
\cmidrule{2-8}
& 512 & Baseline & - & 68.02 & - & 66.552 & - \\
& 512 & PGT & 0.25 & \textcolor{red}{\textbf{69.5}} & \textcolor{olive}{\textbf{+1.48}} &
\textcolor{blue}{\textbf{67.236}} & \textcolor{olive}{\textbf{+0.684}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC & - & 69.04 & +0.18 & 67.064 & +0.268 \\
& 256 & AGC & - & 69.06 & +0.2 & 67.298 & +0.502 \\
& 256 & PGT & 0.25 & \textcolor{red}{\textbf{69.97}} & \textcolor{olive}{\textbf{+1.11}} &
\textcolor{blue}{\textbf{67.814}} & \textcolor{olive}{\textbf{+1.018}} \\
\cmidrule{2-8}
& 256 & Baseline & - & 68.86 & - & 66.796 & - \\
& 256 & GC+PGT & 0.25 & 68.67 & -0.19 & 67.088 & +0.292 \\
& 256 & AGC+PGT & 0.25 & \textcolor{red}{\textbf{70.92}} & \textcolor{olive}{\textbf{+2.06}} &
\textcolor{blue}{\textbf{68.856}} & \textcolor{olive}{\textbf{+2.06}} \\
\end{tabular}
\end{table}

We provide accuracy results for different batch sizes for the non-BN ResNet-18 trained
on ImageNet-1K and also experiment with different methods (GC, AGC, PGT along with their
combinations). Table \ref{tab:wobn_table} shows the findings. To begin, we use a batch
size of 1024. This is only to illustrate that in unnormalized networks, accuracy suffers
significantly when batch sizes are large. Additionally, the training process is often
unstable at this batch size, resulting in frequent layer dropouts (as seen in
\ref{sec:Empi}) and infinite loss values. Thus for all subsequent tests, we use batch
sizes of 512 and 256. At a batch size of 512, invoking PGT with $\alpha=0.25$ improves
the training accuracy baseline by $1.48\%$ and the test accuracy baseline by $0.684\%$,
while at a batch size of 256, the improvement in training and test accuracies are
$1.11\%$ and $1.018\%$ respectively. In comparison, the test accuracy improvements
obtained by GC and AGC at batch size of 256 is much less, at $0.27\%$ and $0.5\%$,
respectively.
